---
title: "Health Data Science Practical"
author: "Bushra Tabassum"
date: "2025-05-15"
output:
  html_document: default
  pdf_document: 
    fig_width: 12
    fig_height: 6
---

```{r, message=FALSE}
library(tidyverse) 
library(datasauRus)
```

**1. Based on the help file, how many rows and how many columns does the `datasaurus_dozen` file have? What are the variables included in the data frame? (this can be hardcoded)**

Ans: 1846 rows, 3 columns variables included in the data frame: dataset, x and y.

```{r}
datasaurus_dozen %>%
  count(dataset)
```

**2. Plot `y` vs. `x` for the `dino` dataset. Then, calculate the correlation coefficient between `x` and `y` for this dataset.**

```{r}
dino_data <- datasaurus_dozen %>%
  filter(dataset == "dino")
```

```{r fig.fullwidth=TRUE}
ggplot(data = dino_data, mapping = aes(x = x, y = y)) +
  geom_point()

```

```{r, eval=FALSE}
dino_data %>%
  summarize(r = cor(x, y))
```

**The correlation coefficient between `x` and `y` for this dataset=** -0.06447185

```         
**3. Plot `y` vs. `x` for the `star` dataset. You can (and should) reuse code we introduced above, just replace the dataset name with the desired dataset. Then, calculate the correlation coefficient between `x` and `y` for this dataset. How does this value compare to the `r` of `dino`?**
```

```{r}
dino_data <- datasaurus_dozen %>%
  filter(dataset == "star")
ggplot(data = dino_data, mapping = aes(x = x, y = y)) +
  geom_point()
dino_data %>%
  summarize(r = cor(x, y))
```

**The correlation coefficient between `x` and `y` for this dataset=**-0.0629611

```{r}
dino_data <- datasaurus_dozen %>%
  filter(dataset == "dino")
ggplot(data = dino_data, mapping = aes(x = x, y = y)) +
  geom_point()
dino_data %>%
  summarize(r = cor(x, y))
```

3.  **\*\*** 4. Plot `y` vs. `x` for the `circle` dataset. You can (and should) reuse code we introduced above, just replace the dataset name with the desired dataset. Then, calculate the correlation coefficient between `x` and `y` for this dataset. How does this value compare to the `r` of `dino`?{r}

```{r}
dino_data <- datasaurus_dozen %>%
  filter(dataset == "circle")
ggplot(data = dino_data, mapping = aes(x = x, y = y)) +
  geom_point()
dino_data %>%
  summarize(r = cor(x, y))
```

```         
The correlation coefficient between x and y for this dataset= -0.0683
```

```{r}
ggplot(datasaurus_dozen, aes(x = x, y = y, color = dataset))+
  geom_point()+
  facet_wrap(~ dataset, ncol = 3) +
  theme(legend.position = "none")

datasaurus_dozen %>%
  group_by(dataset) %>%
  summarize(r = cor(x, y))
```

```         
```

```         
```

```         
```

**5. Finally, let’s plot all datasets at once. In order to do this we will make use of facetting.**

```{r all-viz, eval=FALSE, fig.fullwidth=TRUE, fig.height=20, fig.width=20}
ggplot(datasaurus_dozen, aes(x = x, y = y, color = dataset))+
  geom_point()+
  facet_wrap(~ dataset, ncol = 3) +
  theme(legend.position = "none")
```

```{r first-10-r, eval=FALSE}
datasaurus_dozen %>%
  group_by(dataset) %>%
  summarize(r = cor(x, y))
```
