---
title: "Practical 1"
author: "Bushra Tabassum"
date: "25-05-25"
output:
  html_document: default
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
setwd("~/Desktop/Assignment Health Data Science")
```
```{r}
read_csv("~/Desktop/Assignment Health Data Science/2016_05v2_VitoriaAppointmentData.csv", 
         col_types = "fffTTifllllflf")
```
```{r}
library(dplyr)          # Data Manipulation
```


```{r}
library(readr)          # Data Input
library(tidymodels)     # Data Manipulation
library(lubridate)      # Data Manipulation
library(dplyr)          # Data Manipulation
library(reshape2)       # Data Manipulation
library(caTools)        # Data Manipulation
library(corrplot)       # Data Visualisation
library(ggplot2)        # Data Visualisation
library(viridis)        # Data Visualisation
library(ggthemes)       # Data Visualisation
library(pROC)           # Metrics
library(caret)          # Machine Learning
library(xgboost)        # xgboost model
library(tidyverse)
```

This practical is based on exploratory data analysis and prediction of a dataset derived from a municipal database of healthcare administrative data. This dataset is derived from Vitoria, the capital city of Espírito Santo, Brazil (population 1.8 million) and was freely shared under a creative commons license.

**Generate an rmarkdown report that contains all the necessary code to document and perform: EDA, prediction of no-shows using XGBoost, and an analysis of variable/feature importance using this data set. Ensure your report includes answers to any questions marked in bold. Please submit your report via brightspace as a link to a git repository containing the rmarkdown and compiled/knitted html version of the notebook.**

## Introduction

The Brazilian public health system, known as SUS for Unified Health System in its acronym in Portuguese, is one of the largest health system in the world, representing government investment of more than 9% of GDP. However, its operation is not homogeneous and there are distinct perceptions of quality from citizens in different regions of the country.  Non-attendance of medical appointments contributes a significant additional burden on limited medical resources.  This analysis will try and investigate possible factors behind non-attendance using an administrative database of appointment data from Vitoria, Espírito Santo, Brazil.

The data required is available via the [course website](https://github.com/maguire-lab/health_data_science_research_2025/tree/master/static_files/practicals/lab1_data).

### Understanding the data

**1** Use the data dictionary describe each of the variables/features in the CSV in your report.
**PatientID: Unique identifier for each patient 
**AppointmentID: Unique identifier to each appointment 
**Gender: Patient Gender (limited to Male or Female)
**ScheduledDate: date on which the appointment was scheduled
**AppointmentDate: date of the actual appointment
**Age: Patient age
**Neighbourhood: District of Vitória in which the appointment 
**SocialWelfare: Patient is a recipient of Bolsa Família welfare payments
**Hypertension: Patient previously diagnoised with hypertensio (Boolean)
**Diabetes: Patient previously diagnosed with diabetes (Boolean)
**AlcoholUseDisorder: Patient previously diagnosed with alcohol use disorder (Boolean)
**Disability: Patient previously diagnosed with a disability (severity rated 0-4)
**SMSReceived: At least 1 reminder text sent before appointment (Boolean)
**NoShow: Patient did not attend scheduled appointment (Boolean: Yes/No)

**2** Can you think of 3 hypotheses for why someone may be more likely to miss a medical appointment?
##a.Individuals without reliable transportation may be unable to travel to their medical appointments
##b.Severe weather conditions like snow or heavy rain may prevent patients from safely reaching the clinic
##c.Long distances to healthcare facilities can make it difficult for patients in underserved areas to attend appointments.

**3** Can you provide 3 examples of important contextual information that is missing in this data dictionary and dataset that could impact your analyses e.g., what type of medical appointment does each `AppointmentID` refer to?  

##Appointment type — The dataset doesn’t specify whether each appointment was for routine care, follow-up, or urgent treatment**
##Patient distance to clinic — There’s no information on how far each patient lives from the appointment location.**
##Weather conditions — The dataset lacks weather data on the day of the appointment, which may affect attendance rates.**

## Data Parsing and Cleaning

**4** Modify the following to make it reproducible i.e., downloads the data file directly from version control

```{r}
raw.data <- read_csv("~/Desktop/Assignment Health Data Science/Health-Data-Science/2016_05v2_VitoriaAppointmentData.csv", 
                     col_types = "fffTTifllllflf")
```

Now we need to check data is valid: because we specified col_types and the data parsed without error most of our data seems to at least be formatted as we expect i.e., ages are integers

```{r}
raw.data %>% filter(Age > 110)
```
We can see there are 2 patient's older than 110 which seems suspicious but we can't actually say if this is impossible.

**5** Are there any individuals with impossible ages? If so we can drop this row using `filter` i.e., `data <- data %>% filter(CRITERIA)`
```{r}
raw.data %>% filter(Age < 0 | Age > 110)
```
##There is 1 individual with -1 age and 2 indiuviduals with 115 age.

```{r}
raw.data %>% filter(Age >= 0 & Age <= 110)
```
```{r}
raw.data %>% filter(Age > 110)
raw.data %>% filter(Age < 0)
raw.data <- raw.data %>% filter(Age >= 0 & Age < 110)
```

## Exploratory Data Analysis
First, we should get an idea if the data meets our expectations, there are newborns in the data (`Age==0`) and we wouldn't expect any of these to be diagnosed with Diabetes, Alcohol Use Disorder, and Hypertension (although in theory it could be possible).  We can easily check this:

```{r}
raw.data %>% filter(Age == 0) %>% select(Hypertension, Diabetes, AlcoholUseDisorder) %>% unique()
```

We can also explore things like how many different neighborhoods are there and how many appoints are from each? 

```{r}
count(raw.data, Neighbourhood, sort = TRUE)
```
**6** What is the maximum number of appointments from the same patient?
```{r}
count(raw.data, PatientID, sort = TRUE)
```
##The maximum number of appointments from the same patient is 88.

Let's explore the correlation between variables:

```{r}

# let's define a plotting function
corplot = function(df){
  
  cor_matrix_raw <- round(cor(df),2)
  cor_matrix <- melt(cor_matrix_raw)
  
  
  #Get triangle of the correlation matrix
  #Lower Triangle
  get_lower_tri<-function(cor_matrix_raw){
    cor_matrix_raw[upper.tri(cor_matrix_raw)] <- NA
    return(cor_matrix_raw)
  }
  
  # Upper Triangle
  get_upper_tri <- function(cor_matrix_raw){
    cor_matrix_raw[lower.tri(cor_matrix_raw)]<- NA
    return(cor_matrix_raw)
  }
  
  upper_tri <- get_upper_tri(cor_matrix_raw)
  
  # Melt the correlation matrix
  cor_matrix <- melt(upper_tri, na.rm = TRUE)
  
  # Heatmap Plot
  cor_graph <- ggplot(data = cor_matrix, aes(Var2, Var1, fill = value))+
    geom_tile(color = "white")+
    scale_fill_gradient2(low = "darkorchid", high = "orangered", mid = "grey50", 
                         midpoint = 0, limit = c(-1,1), space = "Lab", 
                         name="Pearson\nCorrelation") +
    theme_minimal()+ 
    theme(axis.text.x = element_text(angle = 45, vjust = 1, 
                                     size = 8, hjust = 1))+
    coord_fixed()+ geom_text(aes(Var2, Var1, label = value), color = "black", size = 2) +
    theme(
      axis.title.x = element_blank(),
      axis.title.y = element_blank(),
      panel.grid.major = element_blank(),
      panel.border = element_blank(),
      panel.background = element_blank(),
      axis.ticks = element_blank())+
      ggtitle("Correlation Heatmap")+
      theme(plot.title = element_text(hjust = 0.5))
  
  cor_graph
}

numeric.data = mutate_all(raw.data, function(x) as.numeric(x))

# Plot Correlation Heatmap
corplot(numeric.data)

```

Correlation heatmaps are useful for identifying linear relationships between variables/features.
In this case, we are particularly interested in relationships between `NoShow` and any specific variables.

**7** Which parameters most strongly correlate with missing appointments (`NoShow`)?

##The variable most strongly correlated with NoShow is SMSReceived (r = 0.13), though the correlation is weak. No other variables show notable correlation with missed appointments.

**8** Are there any other variables which strongly correlate with one another?

##The strongest correlation is actually between PatientID and AppointmentID (r = 0.65).

**9** Do you see any issues with PatientID/AppointmentID being included in this plot? 

##PatientID and AppointmentID are unique identifiers, not meaningful continuous variables, so including them in a correlation plot is misleading. Their correlations with other variables don’t reflect real-world relationships and can be misleading in analysis.

Let's look at some individual variables and their relationship with `NoShow`.

```{r,fig.align="center"}
ggplot(raw.data) + 
  geom_density(aes(x=Age, fill=NoShow), alpha=0.8) + 
  ggtitle("Density of Age by Attendence")
```
There does seem to be a difference in the distribution of ages of people that miss and don't miss appointments.  
However, the shape of this distribution means the actual correlation is near 0 in the heatmap above. This highlights the need to look at individual variables.

Let's take a closer look at age by breaking it into categories.

```{r, fig.align="center"}
raw.data <- raw.data %>% mutate(Age.Range=cut_interval(Age, length=10))

ggplot(raw.data) + 
  geom_bar(aes(x=Age.Range, fill=NoShow)) + 
  ggtitle("Amount of No Show across Age Ranges")

ggplot(raw.data) + 
  geom_bar(aes(x=Age.Range, fill=NoShow), position='fill') + 
  ggtitle("Proportion of No Show across Age Ranges")

```

**10** How could you be misled if you only plotted 1 of these 2 plots of attendance by age group?

##If we only plotted one of the two charts (either count or proportion), we could be misled as only using the count plot might suggest younger and middle-aged groups miss the most appointments — but that could simply be because there are more people in those groups overall.Also, only using the proportion plot shows the rate of missed appointments across age groups, but not how many people are in each group.So, we could be misled by thinking a group has a high no-show issue just because of large numbers (count), when the rate (proportion) is actually low — or vice versa.

The key takeaway from this is that  number of individuals > 90 are very few from plot 1 so probably are very small so unlikely to make much of an impact on the overall distributions. 
However, other patterns do emerge such as 10-20 age group is nearly twice as likely to miss appointments as the 60-70 years old.

Next, we'll have a look at `SMSReceived` variable:

```{r,fig.align="center"}
ggplot(raw.data) + 
  geom_bar(aes(x=SMSReceived, fill=NoShow), alpha=0.8) + 
  ggtitle("Attendance by SMS Received")

ggplot(raw.data) + 
  geom_bar(aes(x=SMSReceived, fill=NoShow), position='fill', alpha=0.8) + 
  ggtitle("Proportion Attendance by SMS Received")
```


**11** From this plot does it look like SMS reminders increase or decrease the chance of someone not attending an appointment? Why might the opposite actually be true (hint: think about biases)? From the proportion plot, it looks like people who received an SMS reminder (TRUE) actually had a slightly higher no-show rate than those who didn’t.

##From the proportion plot, it looks like people who received an SMS reminder (TRUE) actually had a slightly higher no-show rate than those who didn’t.

##This could be misleading due to bias as SMS reminders may have been sent to people already more likely to miss, or other factors (like age or income) may influence both SMS receipt and attendance.Because younger or wealthier patients often have better access to technology (like mobile phones) and may be more engaged with their health care,hence less likely to miss appointments.

**12** Create a similar plot which compares the the density of `NoShow` across the values of disability 

```{r}
ggplot(raw.data) + 
  geom_bar(aes(x=Disability, fill=NoShow), position='fill', alpha=0.8) + 
  ggtitle("Proportion of Attendance by Disability")
```


Now let's look at the neighbourhood data as location can correlate highly with many social determinants of health. 

```{r, fig.align="center"}
ggplot(raw.data) + 
  geom_bar(aes(x=Neighbourhood, fill=NoShow)) + 
  theme(axis.text.x = element_text(angle=45, hjust=1, size=5)) + 
  ggtitle('Attendance by Neighbourhood')


ggplot(raw.data) + 
  geom_bar(aes(x=Neighbourhood, fill=NoShow), position='fill') + 
  theme(axis.text.x = element_text(angle=45, hjust=1, size=5)) + 
  ggtitle('Proportional Attendance by Neighbourhood')
```

Most neighborhoods have similar proportions of no-show but some have much higher and lower rates.

**13** Suggest a reason for differences in attendance rates across neighbourhoods.
##Some neighbourhoods may be more remote or have poor transportation and communication networks, which likely contributes to the differences in attendance rates across neighbourhoods.

Now let's explore the relationship between gender and NoShow.

```{r, fig.align="center"}
ggplot(raw.data) + 
  geom_bar(aes(x=Gender, fill=NoShow))+
  ggtitle("Attendance by Gender")

ggplot(raw.data) + 
  geom_bar(aes(x=Gender, fill=NoShow), position='fill')+
  ggtitle("Proportion of NoShow by Gender")

```

**14** Create a similar plot using `SocialWelfare`

```{r ,fig.align="center"}
#Insert plot
ggplot(raw.data) + 
  geom_bar(aes(x=SocialWelfare, fill=NoShow))+
  ggtitle("SocialWelfare by attendance")

ggplot(raw.data) + 
  geom_bar(aes(x=SocialWelfare, fill=NoShow), position='fill')+
  ggtitle("Proportion SocialWelfare by attendance")

```
Far more exploration could still be done, including dimensionality reduction approaches but although we have found some patterns there is no major/striking patterns on the data as it currently stands.

However, maybe we can generate some new features/variables that more strongly relate to the `NoShow`.

## Feature Engineering

Let's begin by seeing if appointments on any day of the week has more no-show's. Fortunately, the `lubridate` library makes this quite easy!

```{r}
raw.data <- raw.data %>% mutate(AppointmentDay = wday(AppointmentDate, label=TRUE, abbr=TRUE), 
                                 ScheduledDay = wday(ScheduledDate,  label=TRUE, abbr=TRUE))

ggplot(raw.data) +
  geom_bar(aes(x=AppointmentDay, fill=NoShow)) +
  ggtitle("Amount of No Show across Appointment Day") 

ggplot(raw.data) +
  geom_bar(aes(x=AppointmentDay, fill=NoShow), position = 'fill') +
  ggtitle("Proportion of No Show across Appointment Day") 

```
Let's begin by creating a variable called `Lag`, which is the difference between when an appointment was scheduled and the actual appointment.

```{r, fig.align="center"}
raw.data <- raw.data %>% mutate(Lag.days=difftime(AppointmentDate, ScheduledDate, units = "days"),
                                Lag.hours=difftime(AppointmentDate, ScheduledDate, units = "hours"))

ggplot(raw.data) + 
  geom_density(aes(x=Lag.days, fill=NoShow), alpha=0.7)+
  ggtitle("Density of Lag (days) by attendance")
```

**15** Have a look at the values in lag variable, does anything seem odd?
##Yes, something odd stands out in the Lag (days) variable-there are negative lag values, which means some appointments appear to have been scheduled after the actual appointment date, which isn’t logically possible. This suggests data quality issues may have occured due to incorrect date entries or formatting problems — and such records should be reviewed or removed for accurate analysis.

## Predictive Modeling

Let's see how well we can predict NoShow from the data. 

We'll start by preparing the data, followed by splitting it into testing and training set, modeling and finally, evaluating our results. For now we will subsample but please run on full dataset for final execution.


```{r}
### REMOVE SUBSAMPLING FOR FINAL MODEL
data.prep <- raw.data %>% select(-AppointmentID, -PatientID) #%>% sample_n(10000)

set.seed(42)
data.split <- initial_split(data.prep, prop = 0.7)
train  <- training(data.split)
test <- testing(data.split)
```

Let's now set the cross validation parameters, and add classProbs so we can use AUC as a metric for xgboost.

```{r}
fit.control <- trainControl(method="cv",number=3,
                           classProbs = TRUE, summaryFunction = twoClassSummary)
```

**16** Based on the EDA, how well do you think this is going to work?
##Based on the exploratory data analysis (EDA), the model is unlikely to perform very well. Most variables (like age, SMS received, gender, or social welfare) showed weak relationship with NoShow. The correlation values were close to zero, and the density/proportion plots showed overlapping distributions between those who showed up and those who didn't. While a model like XGBoost may still pick up small, complex patterns, the lack of strong predictive features will likely limit its overall performance. Therefore, we can expect moderate accuracy at best, unless more informative variables such as appointment type are added.

Now we can train our XGBoost model
```{r}
xgb.grid <- expand.grid(eta=c(0.05),
                       max_depth=c(4),colsample_bytree=1,
                       subsample=1, nrounds=500, gamma=0, min_child_weight=5)

xgb.model <- train(NoShow ~ .,data=train, method="xgbTree",metric="ROC",
                  tuneGrid=xgb.grid, trControl=fit.control)

xgb.pred <- predict(xgb.model, newdata=test)
xgb.probs <- predict(xgb.model, newdata=test, type="prob")
```

```{r}
test <- test %>% mutate(NoShow.numerical = ifelse(NoShow=="Yes",1,0))
confusionMatrix(xgb.pred, test$NoShow, positive="Yes")
paste("XGBoost Area under ROC Curve: ", round(auc(test$NoShow.numerical, xgb.probs[,2]),3), sep="")
```

This isn't an unreasonable performance, but let's look a bit more carefully at the correct and incorrect predictions,

```{r ,fig.align="center"}
xgb.probs$Actual = test$NoShow.numerical
xgb.probs$ActualClass = test$NoShow
xgb.probs$PredictedClass = xgb.pred
xgb.probs$Match = ifelse(xgb.probs$ActualClass == xgb.probs$PredictedClass,
                         "Correct","Incorrect")
# [4.8] Plot Accuracy
xgb.probs$Match = factor(xgb.probs$Match,levels=c("Incorrect","Correct"))
ggplot(xgb.probs,aes(x=Yes,y=Actual,color=Match))+
  geom_jitter(alpha=0.2,size=0.25)+
  scale_color_manual(values=c("grey40","orangered"))+
  ggtitle("Visualizing Model Performance", "(Dust Plot)")
```


Finally, let's close it off with the variable importance of our model:

```{r,fig.align="center"}
results = data.frame(Feature = rownames(varImp(xgb.model)$importance)[1:10],
                     Importance = varImp(xgb.model)$importance[1:10,])

results$Feature = factor(results$Feature,levels=results$Feature)


# [4.10] Plot Variable Importance
ggplot(results, aes(x=Feature, y=Importance,fill=Importance))+
  geom_bar(stat="identity")+
  scale_fill_gradient(low="grey20",high="orangered")+
  ggtitle("XGBoost Variable Importance")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

**17** Using the [caret package](https://topepo.github.io/caret/) fit and evaluate 1 other ML model on this data.

```{r}
library(caret)
```
```{r}
rf.grid <- expand.grid(mtry = 3) 
```
```{r}
rf.model <- train(NoShow ~ ., 
                  data = train,
                  method = "rf",
                  metric = "ROC",
                  tuneGrid = rf.grid,
                  trControl = fit.control,
                  ntree = 100)
```
```{r}
rf.pred <- predict(rf.model, newdata = test)
rf.probs <- predict(rf.model, newdata = test, type = "prob")
```
```{r}
confusionMatrix(rf.pred, test$NoShow, positive = "Yes")
```
```{r}
paste("Random Forest Area under ROC Curve: ", 
      round(auc(test$NoShow.numerical, rf.probs[,2]), 3), sep = "")
```

##The Random Forest model was fitted using the caret package. It achieved an AUC of 0.72, slightly lower than XGBoost (0.743).

**18** Based on everything, do you think we can trust analyses based on this dataset? Explain your reasoning.

##We can trust the analysis to a point, but with caution. There were odd entries, like negative ages or appointments scheduled after they happened.Also, we don’t know what the appointments were for, how far people had to travel, or what the weather was like—factors that could impact attendance.Most of the variables also didn’t show a strong relationship with whether a person missed their appointmen. Hence, strong conclusions based on this dataset should be made with caution.
## Credits

This notebook was based on a combination of other notebooks e.g., [1](https://www.kaggle.com/code/tsilveira/applying-heatmaps-for-categorical-data-analysis), [2](https://www.kaggle.com/code/samratp/predict-show-noshow-eda-visualization-model), [3](https://www.kaggle.com/code/andrewmvd/exploring-and-predicting-no-shows-with-xgboost/report)